Nail: A practical interface generator for data formats

### 2014-04-02 paper/design.tex
done with pass over section 3 (design), another XXX note

Nail also features \emph{constant parsers}, which are not affecting the internal
representation. Constant parsers can appear instead of structure fields and as
separators in the SepBy combinator.

The simplest constant parser is an integer or array of integers with fixed
value, e.g. \texttt{uint8=0}, \texttt{many uint8=[1,2]}. A convenience notation
for strings is also supported: \texttt{many uint8 = "foo"}.

\paragraph{Wrap combinator.} When implementing real protocols with Nail, we often found that many structures
consist of many constant parsers and only one named field. This pattern is
common in binary protocols which use fixed headers to denote the type of data
structure to be parsed.  In order to keep the internal representation cleaner,
we introduced the wrap combinator, which takes a sequence of parsers containing
exactly one non-constant parser. The parser and generator act as though the wrap
combinator was a sequence of parsers, but the data model does not wrap the
single value in another structure, making the application code more concise.\\
Example: \\ \verb+<uint8='"';many int8|'a'..'z'; uint8='"'>+  parses a quoted
lowercase word into an array - excluding the quotation marks.

\paragraph{Constant combinators.}
In some protocols, there might be many ways to represent the same constant field
and there is no semantic difference between different syntaxes.  Nail therefore
supports choice (\texttt{|| uint8 = 0 || uint16 = 0xFFFF}) and repetition
(\texttt{many (uint8= ' ')}). This is particularly useful for handling padding or whitespace.

@@

\subsection{Dependent fields}
Another problem for parser generators is that binary protocols often contain
length and offset fields. While conventional parsing algorithms can deal with
bounded offset fields - a finite automaton can count a bounded integer, and we
can feed the (finite) input multiple times to the finite automaton. However,
this implementation is both time-inefficient - why feed many bytes into the
automaton that will just be skipped - and very cumbersome to express with
current parser generators. Therefore, if languages with offset fields are parsed
with parser generators, the only currently feasible way is to add 'ad-hoc' hacks
such as changing the input pointer of the generated parser on the fly from
semantic actions.

Nail will properly support both offset and length fields and much of the
following discussion applies to both , although the current prototype only
implements lengths, which we will focus on.

We call length or offset fields \textit{dependent fields}, because during
parsing, another parser depends on them, and while generating output, their
value depends on some other structure in the data model. Dependency fields
appear in a structure combinator as would any other integer field, but their
name begins with an '@' sign. Dependency fields are not exposed in the data
model, but transparently computed.

A dependency field has to appear in the grammar before it can be used. 
\paragraph{Length fields: n_of}
The length combinator takes a dependency field and a parser, evaluating the
parser as often as the value of the dependency field, returning an array of the
values. When generating output, it emits the array and writes its length to the
dependency field.
\paragraph{Offsets: offset} 
The offset combinator takes a dependency field and a parser. It moves the parser
to the position specified in the offset field and invokes the inner parser, and
then moves the input back to its original position.
While generating output, all structures referred to by offset are generated
after the main structure and the dependent offset fields patched up.

\paragraph{Checksums: raw_depend}
In many data formats, some values depend on external representation, such as
checksums and cryptographic signatures. While it would be possible to extend
our constraint language to be powerful enough to support such constructs, we
would essentially be building a separate, Turing complete language that has all
the same pitfalls existing programs have. Therefore, we intend to  allow the programmer to
very carefully escape Nail's programming model and write a function that takes a
pointer to the dependent value and a range of bytes. 
This feature is not implemented in the current prototype. \\
Example:\\
 \verb+{ data many uint8; @checksum uint32+\\
 \verb+raw_depend @checksum data crc32 }+\\
 \verb+/*bool crc32(uint32_t *out, uint8_t *in);*/+
Because this feature compromises Nail's security guarantees, it should only be
used in limited circumstances and with carefully prepared checksum functions.

>>>>

Nail also features \emph{constant parsers}, which do not affect the internal
representation. Constant parsers can appear instead of structure fields and as
separators in the sepBy combinator.

The simplest constant parser is an integer or array of integers with fixed
value; for example, \texttt{uint8=0}, or \texttt{many uint8=[1,2]}. A convenience notation
for strings is also supported: \texttt{many uint8 = "foo"}.

\paragraph{Wrap combinator.} When implementing real protocols with Nail, we often found that
structures that consist of many constant parsers and only one named field. This pattern is
common in binary protocols which use fixed headers to denote the type of data
structure to be parsed.  In order to keep the internal representation cleaner,
we introduced the wrap combinator, which takes a sequence of parsers containing
exactly one non-constant parser. The parser and generator act as though the wrap
combinator was a sequence of parsers, but the data model does not wrap the
single value in another structure, making the application-visible representation
(and thus application code) more concise.

For example, \texttt{<uint8='"'; many int8 | 'a'..'z'; uint8='"'>} parses a quoted
lowercase word into an array, excluding the quotation marks.

\paragraph{Constant combinators.}
In some protocols, there might be many ways to represent the same constant field
and there is no semantic difference between the different syntactic representations.
To support this pattern, Nail allows developers to use choice and repetition
combinators together with constant fields, such as
\texttt{many (uint8=' ')} (representing any number of space characters), or
\texttt{|| uint8 = 0 || uint16 = 0xFFFF} (\XXX[what is this doing?  what is this
syntax?  it looks notably different from Figure~\ref{fig:choice}]).
This is particularly useful for handling padding or whitespace.

@@

\subsection{Dependent fields}
Another problem for parser generators is that binary protocols often contain
length and offset fields. Conventional parsing algorithms can, in principle,
deal with
bounded offset fields: a finite automaton can count a bounded integer, and we
can feed the (finite) input multiple times to the finite automaton. However,
this implementation is both time-inefficient (it feeds many bytes into the
automaton that will just be skipped) and very cumbersome to express with
current parser generators. Therefore, if languages with offset fields need
to be parsed
with parser generators, the only currently practical way is to add ad-hoc hacks
such as changing the input pointer of the generated parser on the fly, as part
of the code in the semantic action for the offset field.

Nail will properly support both offset and length fields and much of the
following discussion applies to both, although the current prototype only
implements lengths, which we will focus on.

We call length or offset fields \textit{dependent fields}, because during
parsing, another parser depends on them, and while generating output, their
value depends on some other structure in the data model. Dependency fields
appear in a structure combinator as would any other integer field, but
their name begins with an \texttt{@} sign.  A dependency field has to
appear in the grammar before it can be used.

Dependency fields are not exposed in the data model, but instead are
transparently computed.  This frees the developer from checking that
these fields are correct (for input) or having to keep their values in
sync with the rest of the data structure (for output).

\paragraph{Length fields.}

The length combinator, \texttt{n_of}, takes a dependency field $n$
and a parser, evaluates the parser exactly $n$ times (i.e., setting
the number of iterations to be the $n$ field's value), and returns
an array of the parser's values. When generating output, it emits
the array and writes its length to the dependency field.

\paragraph{Offsets.}

The \texttt{offset} combinator takes a dependency field and a parser. It
moves the parser to the position specified in the offset field and
invokes the inner parser, and then moves the input back to its original
position.  While generating output, all structures referred to by offset
are generated after the main structure and the dependent offset fields
are patched up.

\paragraph{Checksums.}

In many data formats, some values depend on external representation,
such as checksums and cryptographic signatures.  While it would be
possible to extend our constraint language to be powerful enough to
support such constructs, we would essentially be building a separate,
Turing-complete language that has all the same pitfalls existing programs
have.  Therefore, we intend to allow the programmer to carefully escape
Nail's programming model and write a function that takes a pointer to
the dependent value and a range of bytes, using the \texttt{raw_depend}
combinator.

For example, we imagine the following grammar could be used to represent
a sequence of bytes \texttt{data} followed by its CRC32 checksum:

\begin{verbatim}
data many uint8; @checksum uint32
raw_depend @checksum data crc32
\end{verbatim}

\noindent
where \texttt{crc32} is a function supplied by the application, with
the following signature:

\begin{verbatim}
bool crc32(uint32_t *out, uint8_t *in);
\end{verbatim}

Because this feature compromises Nail's security guarantees, it should
only be used in limited circumstances and with carefully prepared checksum
functions.  This feature is not implemented in the current prototype.
